OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-03-05 paper/paper.Rmd
Cite everybody

- **TODO: summarize results & conclusion**

@@

- **TODO: ask Sarah Westcott to check the accuracy of this description**

@@

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. There are two strategies for
generating OTUs with OptiFit: 1) fit sequences to reference OTUs of an
independent database, or 2) split the dataset into a reference and query
fraction, cluster the reference sequences _de novo_, then fit the query
sequences to the reference OTUs. For each dataset repeated with 100 random
seeds, we generated OTUs with OptiFit using both strategies. To compare to
existing software, we also clustered OTUs _de novo_ using OptiClust and vsearch,
and with vsearch in reference-based mode against the greengenes database. All
clustering was performed at a distance threshold of 0.03 and OTU quality was
evaluated using the Matthews Correlation Coefficient (MCC) as described
previously [@westcott_opticlust_2017]. We calculated the fraction of query
sequences that were fit to existing OTUs in closed reference mode as an
additional measure of quality for this mode. _De novo_ clustering with OptiFit
produced **X%** higher quality OTUs than vsearch, but performed **Y%** slower
than vsearch.

@@

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database (v13_8_99), Silva non-redundant database (v132), or the Ribosomal
Database Project (RDP; v16). **TODO: cite database papers?** In open reference
mode, fitting the datasets to reference OTUs with OptiFit produced OTUs of
similar quality as clustering the datasets _de novo_ with OptiClust (**TODO:
diff. in medians**). This held true for all datasets and reference databases.
OptiFit produced higher quality OTUs than vsearch when open reference clustering
against the greengenes database, with median MCC scores of **A** and **B%**
(respectively). However, open reference vsearch and OptiClust respectively ran
**X%** and **Y%** faster than OptiFit in open reference mode.

In closed reference mode, OptiFit produced lower quality OTUs than OptiClust by
**X%** when fitting sequences to Greengenes and Silva, and **Y%** worse when
fitting to RDP. Only up to **Z%** of query sequences were fit to reference OTUs
in closed reference mode across any dataset/database combination. OptiFit was
able to fit **X** more query sequences to reference OTUs created with the
Greengenes and Silva databases than with RDP. vsearch was able to fit **W%**
more query sequences to the greengenes reference than OptiFit in closed
reference mode. In terms of run time, closed reference OptiFit outperformed
OptiClust by **A%** and vsearch by **B%**.

@@

A split dataset strategy was employed to assess how well OptiFit performs for
tasks where new sequences are added to existing OTUs, such as when comparing OTUs
across studies or making predictions with machine learning models.
Datasets were randomly split into a reference fraction and a query fraction.
Reference sizes from 10% to 90% of the sequences were created, with the
remaining sequences used for the query. Reference sequences were clustered _de
novo_ with OptiClust, then query sequences were fit to the _de novo_ OTUs with
OptiFit.

@@

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP, which has only about 16,000 V4
16S sequences compared to about 170,000 in Silva and greengenes. We recommend
users who require an independent reference database opt for large databases with
good coverage of microbial diversity. Since OptiClust performs faster than open
reference OptiFit and creates higher quality OTUs than closed reference OptiFit
with the database strategy, we recommend using OptiClust rather than fitting to
a database whenever stable OTUs are not required for the study at hand.

The mothur algorithms produced higher quality OTUs than vsearch in open
reference, closed reference, or _de novo_ modes. However, vsearch was able to
fit more sequence into OTUs than OptiFit in closed reference mode. While both
mothur and vsearch use a distance or similarity threshold for determining how to
assign sequences to OTUs, vsearch is more permissive than mothur. Mothur's
OptiFit and OptiClust use all of the sequences to define an OTU, requiring that
all pairs of sequences (including reference and query sequences) in an OTU are
within the distance threshold without penalizing the MCC. In contrast, vsearch
only requires each query sequence to be similar to the single sequence that
seeded the OTU. In this way, vsearch sacrifices OTU quality in order to allow
more sequences to fit to OTUs. Users who require closed reference clustering to
the greengenes database may prefer to use vsearch if they wish to maximize the
fraction of sequences that can be fit at the cost of OTU quality. However,
mothur's OptiClust or OptiFit are recommended for _de novo_ or open reference
clustering to produce OTUs of the highest possible quality.

@@

@@

We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] and
relied on Python [@van_rossum_python_2009], R [@r_core_team_r_2020], and GNU
bash. Software used includes mothur v1.45.0 [@schloss_introducing_2009], vsearch
v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the Tidyverse
metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], and the conda
environment manager [@noauthor_anaconda_2016]. The complete workflow,
manuscript, and conda environment are available at **TODO: UPDATED REPO LINK**.

>>>>

**TODO: summarize results & conclusion**

@@

**TODO: ask Sarah Westcott to check the accuracy of this description**

@@

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. There are two strategies for
generating OTUs with OptiFit: 1) fit sequences to reference OTUs of an
independent database, or 2) split the dataset into a reference and query
fraction, cluster the reference sequences _de novo_, then fit the query
sequences to the reference OTUs. For each dataset repeated with 100 random
seeds, we generated OTUs with OptiFit using both strategies. To compare to
existing software, we also clustered OTUs _de novo_ using OptiClust and vsearch,
and with vsearch in reference-based mode against the Greengenes database. All
clustering was performed at a distance threshold of 0.03 and OTU quality was
evaluated using the Matthews Correlation Coefficient (MCC) as described
previously [@westcott_opticlust_2017]. We calculated the fraction of query
sequences that were fit to existing OTUs in closed reference mode as an
additional measure of quality for this mode. _De novo_ clustering with OptiClust
produced **X%** higher quality OTUs than vsearch, but performed **Y%** slower
than vsearch.

@@

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database, SILVA non-redundant database, or the Ribosomal Database Project (RDP)
[@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].

In open reference mode, fitting the datasets to reference OTUs with OptiFit
produced OTUs of similar quality as clustering the datasets _de novo_ with
OptiClust (**diff. in medians**).
This held true for all datasets and reference databases.
OptiFit produced higher quality OTUs than vsearch when open reference clustering
against the Greengenes database, with median MCC scores of **A** and **B%**
(respectively).
However, open reference vsearch and OptiClust respectively ran **X%** and **Y%**
faster than OptiFit in open reference mode.

In closed reference mode, OptiFit produced lower quality OTUs than OptiClust by
**X%** when fitting sequences to Greengenes and SILVA, and **Y%** worse when
fitting to RDP. Only up to **Z%** of query sequences were fit to reference OTUs
in closed reference mode across any dataset/database combination. OptiFit was
able to fit **X** more query sequences to reference OTUs created with the
Greengenes and SILVA databases than with RDP. vsearch was able to fit **W%**
more query sequences to the Greengenes reference than OptiFit in closed
reference mode. In terms of run time, closed reference OptiFit outperformed
OptiClust by **A%** and vsearch by **B%**.

@@

A split dataset strategy was employed to assess how well OptiFit performs for
tasks where new sequences are added to existing OTUs, such as when comparing OTUs
across studies or making predictions with machine learning models.
Datasets were randomly split into a reference fraction and a query fraction.
Reference sizes from 10% to 80% of the sequences were created, with the
remaining sequences used for the query. Reference sequences were clustered _de
novo_ with OptiClust, then query sequences were fit to the _de novo_ OTUs with
OptiFit.

@@

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP, which has only about 16,000 V4
16S sequences compared to about 170,000 in SILVA and Greengenes. We recommend
users who require an independent reference database opt for large databases with
good coverage of microbial diversity. Since OptiClust performs faster than open
reference OptiFit and creates higher quality OTUs than closed reference OptiFit
with the database strategy, we recommend using OptiClust rather than fitting to
a database whenever stable OTUs are not required for the study at hand.

The mothur algorithms produced higher quality OTUs than vsearch in open
reference, closed reference, or _de novo_ modes. However, vsearch was able to
fit more sequence into OTUs than OptiFit in closed reference mode. While both
mothur and vsearch use a distance or similarity threshold for determining how to
assign sequences to OTUs, vsearch is more permissive than mothur. Mothur's
OptiFit and OptiClust use all of the sequences to define an OTU, requiring that
all pairs of sequences (including reference and query sequences) in an OTU are
within the distance threshold without penalizing the MCC. In contrast, vsearch
only requires each query sequence to be similar to the single sequence that
seeded the OTU. In this way, vsearch sacrifices OTU quality in order to allow
more sequences to fit to OTUs. Users who require closed reference clustering to
the Greengenes database may prefer to use vsearch if they wish to maximize the
fraction of sequences that can be fit at the cost of OTU quality. However,
mothur's OptiClust or OptiFit are recommended for _de novo_ or open reference
clustering to produce OTUs of the highest possible quality.

@@

We downloaded 16S rRNA gene amplicon sequences from four published datasets
isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples.
Raw sequences were processed using mothur according to the Schloss Lab MiSeq SOP
as described in the mothur wiki and accompanying study by Kozich _et al._
[@schloss_miseq_nodate; @kozich_development_2013].
These steps included trimming and filtering for quality, aligning to the SILVA
reference alignment [@quast_silva_2013], discarding sequences that aligned
outside the V4 region, removing chimeric reads with UCHIME [@edgar_uchime_2011],
and calculating distances between all pairs of sequences within each dataset
prior to clustering.

### Reference database clustering

To generate reference OTUs from independent databases, we downloaded sequences
from the Greengenes database (v13_8_99) [@desantis_greengenes_2006], SILVA
non-redundant database (v132) [@quast_silva_2013], and the Ribosomal Database
Project (v16) [@cole_ribosomal_2014].
These sequences were processed using the same steps outlined above followed by
clustering sequences into _de novo_ OTUs with OptiClust.
Processed reads from each of the four datasets were clustered with OptiFit to
the reference OTUs generated from each of the three databases.
When reference clustering with vsearch, processed datasets were fit directly to
the unprocessed Greengenes reference alignment, since this method is how vsearch
is typically used by the QIIME2 software reference-based clustering
[@bolyen_reproducible_2019; @noauthor_clustering_nodate].

### Split dataset clustering

**TODO: describe split dataset creation**

@@

We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] and
wrote scripts in R [@r_core_team_r_2020], Python [@van_rossum_python_2009], and
GNU bash [@noauthor_bash_nodate].
Software used includes mothur v1.45.0 [@schloss_introducing_2009], vsearch
v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the Tidyverse
metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], the SRA toolkit
[@noauthor_sra-tools_nodate], and the conda environment manager
[@noauthor_anaconda_2016].
The complete workflow, manuscript, and conda environment are available at
**TODO: UPDATED REPO LINK**.
